---
title: "Bayesian data analysis & cognitive modeling"
subtitle: "03: probability & statistics primer"
author: "Michael Franke"
output:
  ioslides_presentation:
    css: mistyle.css
    smaller: no
    transition: faster
    widescreen: yes
---

```{r setup, include=FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, dev.args = list(bg = 'transparent'), fig.align='center')
require('tidyverse')
require('forcats')
theme_set(theme_bw() + theme(plot.background=element_blank()) )
```

<script type="text/x-mathjax-config">
  MathJax.Hub.Config({ TeX: { extensions: ["color.js"] }});
</script>

```{r, child = "miincludes.Rmd"}

```




## road map for today

<span style = "color:white"> &nbsp; </span>

- probability
    - discrete, continuous, cumulative
    - subjective vs. objective chance
    - means, modes, medians
    
<span style = "color:white"> &nbsp; </span>

- conditional probability & Bayes rule

<span style = "color:white"> &nbsp; </span>

- $p$-values & confidence intervals

# probability

## definition

a <span style = "color:firebrick">discrete probability distribution</span> over a finite set of mutually exclusive world states $\States$ is a function $P \colon \pow{\States} \rightarrow [0;1]$ such that:
    
- $P(\States') = \sum_{\state \in \States'} P(\state)$, for all $\States' \subseteq \States$
    - here $P(\state)$ is shorthand for $P(\set{\state})$
- $P(\States)=1$

## example

okay, winter is coming; but who will sit the Iron Throne next spring?

<span style = "color:white"> &nbsp; </span>

```{r}
house.probs = c(6,3,1,2,4) %>% (function(x) x/sum(x))
names(house.probs)= c("Targaryen", "Lannister", "Baratheon", "Greyjoy", "Stark")
round(house.probs,3)
```

<span style = "color:white"> &nbsp; </span>

```{r}
sum(house.probs)
```


## example

```{r, fig.height = 3.5}
house.probs.df = as_tibble(house.probs) %>% 
  mutate(house = names(house.probs) %>% factor() %>% fct_inorder()) %>% 
  rename(probability = value)
house.plot = ggplot(house.probs.df, aes(x = house, y = probability)) + 
  geom_bar(stat = "identity", fill = "firebrick")
house.plot
```

## notation

<span style = "color:white"> &nbsp; </span>

if $f \colon \States \rightarrow \mathbb{R}^{\ge0}$, then

$$ P(\state) \propto f(\state) $$

is shorthand notation for

$$ P(\state) = \frac{f(\state)}{ \sum_{\state' \in \States} f(\state')} $$

<span style = "color:white"> &nbsp; </span>

<span style = "color:firebrick">example</span>

```{r, eval = F}
house.probs = c(6,3,1,2,4) %>% (function(x) x/sum(x))
```


# fini

## to prevent boredom

<span style = "color:firebrick">obligatory</span>

- prepare Kruschke Chapter 4

- fill in [CSV file](https://github.com/michael-franke/BDACM_2017/blob/master/data/01_class_enquette.csv) and send it to [Christian Adam](mailto:c.adam@student.uni-tuebingen.de)

<span style = "color:white"> &nbsp; </span>

<span style = "color:firebrick">optional</span>

- read more on [R for Data Science](http://r4ds.had.co.nz)
    - Chapters 5 and 12 on data representation and wrangling

- explore reading times in today's data set

<span style = "color:white"> &nbsp; </span>

