---
title: "Bayesian data analysis & cognitive modeling"
subtitle: "07: JAGS"
author: "Michael Franke"
output:
  ioslides_presentation:
    css: mistyle.css
    transition: faster
    widescreen: yes
---

```{r setup, include=FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, 
                      dev.args = list(bg = 'transparent'), fig.align='center',
                      cache=FALSE)
require('tidyverse')
require('forcats')
theme_set(theme_bw() + theme(plot.background=element_blank()) )
```

<script type="text/x-mathjax-config">
  MathJax.Hub.Config({ TeX: { extensions: ["color.js"] }});
</script>

<script type="text/x-mathjax-config">
MathJax.Hub.Register.StartupHook("TeX Jax Ready",function () {
  var MML = MathJax.ElementJax.mml,
      TEX = MathJax.InputJax.TeX;

  TEX.Definitions.macros.bfrac = "myBevelFraction";

  TEX.Parse.Augment({
    myBevelFraction: function (name) {
      var num = this.ParseArg(name),
          den = this.ParseArg(name);
      this.Push(MML.mfrac(num,den).With({bevelled: true}));
    }
  });
});
</script>


```{r, child = "miincludes.Rmd"}

```

## roadmap

- JAGS
    - background
    - model specification syntax
    - workflow
    - tips & tricks

## recap

Bayes rule for data analysis:

$$\underbrace{P(\theta \, | \, D)}_{posterior} \propto \underbrace{P(\theta)}_{prior} \times \underbrace{P(D \, | \, \theta)}_{likelihood}$$

normalizing constant:

$$ \int P(\theta') \times P(D \mid \theta') \, \text{d}\theta' = P(D) $$

easy to solve only if:

- $\theta$ is a single discrete variable with reasonably sized domain
- $P(\theta)$ is conjugate prior for the likelihood function $P(D \mid \theta)$
- we are very lucky

## recap 

<span style = "color:firebrick">Markov Chain Monte Carlo</span>

get sequence of samples $x_1, \dots, x_n$ s.t.

1. sequence has the <span style = "color:firebrick">Markov property</span> ($x_{i+1}$ depends only on $x_i$), and
2. the <span style = "color:firebrick">stationary distribution</span> of the chain is $P$.

<span style = "color:firebrick">MCMC algorithms</span>

- Metropolis Hastings
    - versatile but often inefficient
    - depends on proposal distribution
- Gibbs sampling
    - fast and efficient, but not universally applicable
    - depends on availability of conditional posterior
    
## recap     

<span style = "color:firebrick">assessing quality of sample chains</span>

- convergence / representativeness
    - trace plots
    - $\hat{R}$
- efficiency
    - autocorrelation
    - effective sample size

# JAGS

## history

<div style = "float:left; width:45%;">

BUGS project (1989 - present ???)

- Bayesian inference Using Gibbs Sampling
- developed by UK-based biostatisticians

<span style = "color:white"> &nbsp; </span>


[WinBUGS](http://www.mrc-bsu.cam.ac.uk/software/bugs/) (1997 - 2007)

- Windows based; with GUI
- component pascal

  
</div>
<div style = "float:right; width:45%;">

[OpenBUGS](http://openbugs.net/w/FrontPage) (2005 - present) 

- Windows & Linux; MacOS through [Wine](https://www.winehq.org)
- component pascal

<span style = "color:white"> &nbsp; </span>



[JAGS](http://mcmc-jags.sourceforge.net) (2007 - present)

- Windows, Linux, MacOS
- no GUI
- written in C++
  
</div>  


## JAGS

<span style = "color:firebrick">Just Another Gibbs Sampler</span>

- declarative language
    - model as a directed acyclic graph
    - nodes are variables, edges are deterministic or probabilistic dependencies
- automatically selects adequate sampler
- call from & process output in R via packages:
    - `R2Jags`, `rjags` ...

<div style = "position:absolute; top: 620px; right:60px;">
  coded by [Martyn Plummer](https://www.iarc.fr/en/staffdirectory/displaystaff.php?id=10118)
</div>


## example

```{r, results='hide', warning=FALSE, message=FALSE}
require('rjags') # also loads `coda` package
modelString = "
model{
  theta ~ dbeta(1,1)
  k ~ dbinom(theta, N)
}"
# prepare for JAGS
dataList = list(k = 7, N = 24)
# set up and run model
jagsModel = jags.model(file = textConnection(modelString), 
                       data = dataList,
                       n.chains = 2)
update(jagsModel, n.iter = 5000)
codaSamples = coda.samples(jagsModel, 
                           variable.names = c("theta"),
                           n.iter = 5000)
```

## example

```{r}
ms = ggmcmc::ggs(codaSamples)
ms %>% group_by(Parameter) %>% 
   summarise(mean = mean(value),
             HDIlow  = coda::HPDinterval(as.mcmc(value))[1], 
             HDIhigh = coda::HPDinterval(as.mcmc(value))[2])
```

## example

```{r, fig.align='center', fig.width=7, fig.height=3.5}
  tracePlot = ggmcmc::ggs_traceplot(ms)  
  densPlot = ggmcmc::ggs_density(ms) + 
    stat_function(fun = function(x) dbeta(x, 8,18), color = "black")
```

```{r, echo = FALSE}
  tracePlot = tracePlot +
    theme(plot.background=element_blank())
  densPlot = densPlot + 
    theme(plot.background=element_blank())
```

```{r, fig.align='center', fig.width=7, fig.height=3.5}
  gridExtra::grid.arrange(tracePlot, densPlot, ncol = 2) 
```

# Using JAGS

## about JAGS

- current version 4
    - faster, allows for use of `=`
    - added samplers and distributions (not yet all documented!?)
- syntax is a mix of BUGS and R
    - careful with parameterization of probability distributions !!!
- declarative model specification
    - directed acyclic graphs / Bayes nets
        - nodes are variables, edges are dependencies
    - order invariant
    - no variable reassignment, no control flow statements etc.
        - exception: `ifelse`, `step` ... commands for boolean switching
        - `for` loops for vector construction
- $\exists$ command-line interface (not used in this course)

## structure of a JAGS model description

<span style = "color:white"> &nbsp; </span>


```{r, eval = FALSE}
# declare size of variables if needed
var ... ; 
# do some data massaging (usually done in R)
data{
  ...
}
# specify model
model{
  ...
}
```


<span style = "color:white"> &nbsp; </span>

- you could write this into a separtate file called "myModel.jags.R"
    - this is untidy file naming but gives you default R syntax highlighting

## (nonsense) example

```{r, eval = FALSE}

var myData[5,100], myMu[5], myTau[5]; 
data{
  N = sum(counts) # counts is input to JAGS from R
  indVector = c(3,2,4) # works like in R
  specialConditions = counts[indVector] # like R; new in JAGS 4
}
model{
  for (i in 1:dim(myData)[1]){
    for (j in 1:dim(myData)[2]){
      myData[i,j] ~ dnorm(myMu[i], myTau[i])
    }
  }
  for (i in 1:5){
    tmp[i] ~ dbeta(1,1)
    myMu[i] = 100 + tmp[i] - 0.5
    myTau[i] ~ dunif(0, 1000)
  }
}
```


## caveats

- semicolon after declaration of variable dimensions
- declare dimensions to avoid confusing the JAGS compiler
- JAGS error messages are occassionally underinformative
- all model specification lines are probabilistic ("~") or deterministic ("=" or "<-")
    - "=" only works for JAGS 4+
- all probabilistic dependencies must be samples from a distribution known to JAGS!
    - this is therefore all ruled out:
    
```{r, eval = FALSE}

  x ~ dbeta(1,1) - 0.5

  myCostumDistribution = ... # something fancy
  x ~ myCustomDistribution(0,1)
  
```
    
## another example    
   
Your turn: what's this model good for?

- `obs` is a vector of 100 observations (real numbers)

```{r, eval = FALSE}
model{
  mu ~ dunif(-2,2)
  var ~ dunif(0, 100)
  tau = 1/var
  for (i in 1:100){
    obs[i] ~ dnorm(mu,tau)
  }
}
``` 

NB: JAGS uses precision $\tau = 1/\sigma^2$, not standard deviation $\sigma$ in `dnorm`

## model specifications, formally

```{r, eval = FALSE}
model{
  mu ~ dunif(-2,2)
  var ~ dunif(0, 100)
  tau = 1/var
  for (i in 1:100){
    obs[i] ~ dnorm(mu,tau)
  }
}
``` 

$$\mu \sim \text{Unif}(-2,2)$$
$$\sigma^2 \sim \text{Unif}(0,100)$$
$$obs_i \sim \text{Norm}(\mu, \sigma)$$
    
# running a JAGS script    



```{r, echo = FALSE}
require('coda')

MH = function(f, iterations = 50, chains = 2, burnIn = 0){
  out = array(0, dim = c(chains, iterations - burnIn, 2))
  dimnames(out) = list("chain" = 1:chains, 
                       "iteration" = 1:(iterations-burnIn), 
                       "variable" = c("mu", "sigma"))
  for (c in 1:chains){
    mu = runif(1, min = -2, max = 2)
    sigma = runif(1, min = 0, max = 4)
    for (i in 1:iterations){
      muNext = mu + runif(1, min = -1.25, max = 1.25)
      sigmaNext = sigma + runif(1, min = -0.25, max = 0.25)
      rndm = runif(1, 0, 1)
      if (f(mu, sigma) < f(muNext, sigmaNext) | f(muNext, sigmaNext) >= f(mu, sigma) * rndm) {
        mu = muNext
        sigma = sigmaNext
      }
      if (i >= burnIn){
        out[c,i-burnIn,1] = mu
        out[c,i-burnIn,2] = sigma
      }
    }
  }
  return(mcmc.list(mcmc(out[1,,]), mcmc(out[2,,])))
}
```

## homebrew MH samples

```{r, fig.align='center', fig.width=7, fig.height=3.5}
set.seed(1789)

fakeData = rnorm(200, mean = 0, sd = 1)

f = function(mu, sigma){
  if (sigma <=0){
    return(0)
  }
  priorMu = dunif(mu, min = -4, max = 4)
  priorSigma = dunif(sigma, min = 0, max = 4)
  likelihood =  prod(dnorm(fakeData, mean = mu, sd = sigma))
  return(priorMu * priorSigma * likelihood)
}

samplesMH = MH(f, 
               iterations = 60000,
               chains = 2,
               burnIn = 10000) # outputs mcmc.list from `coda` package
```



## back to the future

```{r, results = 'hide'}
modelString = "
model{
  mu ~ dunif(-4,4)
  sigma ~ dunif(0,4)
  tau = 1/sigma^2
  for (i in 1:length(obs)){
    obs[i] ~ dnorm(mu,tau)
  }
}"
jagsModel = jags.model(file = textConnection(modelString), 
                       data = list(obs = fakeData),
                       n.chains = 2)
update(jagsModel, n.iter = 5000)
samplesJAGS = coda.samples(jagsModel, 
                           variable.names = c("mu", "sigma"),
                           n.iter = 5000)
```


## compare sample outputs

```{r, echo = FALSE}
library('ggmcmc')
library('gridExtra')
```


```{r, fig.align='center', fig.width=7, fig.height=4.5}
grid.arrange(ggs_density(ggs(samplesMH)) ,       ggs_density(ggs(samplesJAGS)) )
```

## compare sample outputs

```{r, fig.align='center', fig.width=7, fig.height=4.5}
grid.arrange(ggs_traceplot(ggs(samplesMH)) + theme(plot.background=element_blank()),
      ggs_traceplot(ggs(samplesJAGS)) + theme(plot.background=element_blank()))
```

## compare sample outputs

```{r, echo = FALSE}
DbdaAcfPlot = function( codaObject , parName=varnames(codaObject)[1] , plColors=NULL ) {
  if ( all( parName != varnames(codaObject) ) ) { 
    stop("parName must be a column name of coda object")
  }
  nChain = length(codaObject)
  if ( is.null(plColors) ) plColors=1:nChain
  xMat = NULL
  yMat = NULL
  for ( cIdx in 1:nChain ) {
    acfInfo = acf(codaObject[,c(parName)][[cIdx]],plot=FALSE) 
    xMat = cbind(xMat,acfInfo$lag)
    yMat = cbind(yMat,acfInfo$acf)
  }
  matplot( xMat , yMat , type="o" , pch=20 , col=plColors , ylim=c(0,1) ,
           main="" , xlab="Lag" , ylab="Autocorrelation" )
  abline(h=0,lty="dashed")
  EffChnLngth = effectiveSize(codaObject[,c(parName)])
  text( x=max(xMat) , y=max(yMat) , adj=c(1.0,1.0) , cex=1.25 ,
        labels=paste("ESS =",round(EffChnLngth,1)) )
}
```


```{r, fig.align='center', fig.width=7, fig.height=5}
# function from Kruschke, defined in 'DBDA2E-utilities.R' 
DbdaAcfPlot(samplesMH) + theme(plot.background=element_blank())
```

## compare sample outputs

```{r, fig.align='center', fig.width=7, fig.height=5}
# function from Kruschke, defined in 'DBDA2E-utilities.R'
DbdaAcfPlot(samplesJAGS) + theme(plot.background=element_blank())
```


# tips and tricks

## debugging

develop step by step & monitor each new intermediate variable

```{r, results = 'hide', fig.align='center', fig.width=4, fig.height=2.5}
modelString = "
model{
  mu ~ dnorm(0,1)
}"
jagsModel = jags.model(file = textConnection(modelString), 
                       data = list(obs = fakeData),
                       n.chains = 2, n.adapt = 10)
update(jagsModel, n.iter = 10)
codaSamples = coda.samples(jagsModel, variable.names = c("mu"), n.iter = 5000)
ggs_density(ggs(codaSamples))
```

## prior and posterior predictive

```{r, eval = FALSE}

model{
  thetaPost ~ beta(1,1)
  thetaPrior ~ beta(1,1)
  # generate data from prior distribution
  priorPredictive ~ dbin(thetaPrior,n) 
  # here `thetaPost` is conditioned on observed data!!
  kObs ~ dbin(thetaPost, n) 
  # generate data from posterior
  posteriorPredictive ~ dbin(thetaPost, n) 
}
```

<div style = "float:left; width:35%;">

<span style = "color:firebrick">prior predictive</span>

$$ P(D) = \int P(\theta) \ P(D \mid \theta) \ \text{d}\theta $$

<span style = "color:white"> &nbsp; </span>

</div>
<div style = "float:right; width:55%;">

<span style = "color:firebrick">posterior predictive</span>

$$ P(D \mid D') = \int P(\theta \mid D') \ P(D \mid \theta) \ \text{d}\theta $$

<span style = "color:white"> &nbsp; </span>


</div>  


## conditioning

- boolean operations `x <= y`, `x != y`, `x || y` as usual (see manual)
- functions `ifelse`, `equals` and `step` for boolean conditioning

```{r, eval = FALSE}
model{
  flag ~ dbern(0.5)
  parameter1 = ifelse(flag, 0, -100)
  parameter2 = ifelse(flag, 1, 100)
}
```

## a problem

What is this model trying to achieve? And, why does it not work?

```{r, eval = FALSE}
model{
  flag ~ dbern(0.5)
  SOMEPDF = ifelse(flag, dnorm, dunif)
  parameter1 = ifelse(flag, 0, -100)
  parameter2 = ifelse(flag, 1, 100)
  for(i in 1:length(obs)){
    obs[i] ~ SOMEPDF(parameter1, parameter2)
  }
}
```

## solution

```{r, eval = FALSE}
data{
  for (i in 1:length(obs)){
    ones[i] = 1 # create a vector of ones
  }
}
model{
  flag ~ dbern(0.5)
  parameter1 = ifelse(flag, 0, -100)
  parameter2 = ifelse(flag, 1, 100)
  for(i in 1:length(obs)){
    theta[i] = ifelse(flag, 
                      dnorm(obs[i],parameter1, parameter2), 
                      dunif(obs[i],parameter1, parameter2))
    ones[i] ~ dbern(theta[i])
  }
}
```


# fini

## outlook

<span style = "color:white"> &nbsp; </span>

<span style = "color:firebrick">Friday</span>

- 1st practice session


<span style = "color:white"> &nbsp; </span>

<span style = "color:firebrick">Tuesday</span>

- more complex (interesting) models

## next class ::: we need you prepared!!

<span style = "color:white"> &nbsp; </span>

<span style = "color:firebrick">totally obligatory</span>

- have installed on your laptop, newest versions of:
    - `R` and `JAGS` (optional and recommended: `RStudio`)
    - packages: `R2jags`, `rjags` and `runjags`
    
- obtain access to chapters 3 & 4 of [Lee & Wagenmakers](https://bayesmodels.com) (university library!)
    - download the [example code for JAGS](https://webfiles.uci.edu/mdlee/Code.zip)
    - make sure that you can execute the file `Code/ParameterEstimation/Binomial/Rate_1_jags.R`
        - hint: you may need to plan a rendevous with line 10 of this file

